Modelling state-switching functional data with hidden Markov models

### 2020-10-30 Report/Sections/simulation.tex
commit before adding t_tilde

%\section{Simulation Study}

To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data has equal numbers of short and long dives in expectation, the number subdive states is set to $N^*=2$, and $\Ztwo_{t,t^*}$ is set to a scalar. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which each represent a simulated killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\theta$ are
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
For each dive and subdive segment, we simulate the fine-scale transformed observations $Z^*_t \equiv \left\{Z^*_{t,1}, \ldots, \right\}$, where $Z^*_{t,t^*} = \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$, using the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
Recall that $\Zone_{t,t^*}$ follows a CarHMM and is Normally distributed while $\Ztwo_{t,t^*}$ follows an HMM and is Gamma distributed. It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.

The two simulated dive types differ in that dives of type 1 are much shorter on average (20 seconds) than dives of type 2 (80 seconds). The two subdive types differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher in subdive state 2 than subdive state 1. This corresponds to much more vigorous and variable periodic behaviour in the acceleration data.

All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) are fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

The full CarHHMM-DFT is the best-performing model of the four candidates, which is to be expected as the generating model. If $X^*_{t,t^*}$ is the unknown subdive state and $x^*_{t,t^*}$ is the known true subdive state for dive $t$ and window $t^*$, then using the forward-backward algorithm we find that $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ is $(1 - 10^{-12})$ on average across all simulations, dives, and windows. We refer to this quantity as the average accuracy of the model. For dive type, this average accuracy was above 0.9 (see Table \ref{table:accuracy}). All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases. In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are no more severe for the CarHHMM-DFT than the other three models. The empirical standard error of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard error. See Tables 6-9 of the supplementary material for more information.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its average accuracy and parameter bias. However, the HHMM-DFT differs from the CarHHMM-DFT in its estimates of $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$, both of which are greatly overestimated. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material). These results imply that estimated standard errors of parameter estimates are too small when using the HHMM-DFT since auto-correlation is ignored. This finding is consistent with the results from the case study, where the HHMM-DFT produced higher estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.

The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average accuracy below $0.9$ for both dive and subdive types. This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct observation when determining subdive type. The CarHHMM is also much more likely to misclassify subdive type 1 than subdive type 2 (see Table \ref{table:accuracy}). In addition to its poor average accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters, with the fine-scale probability transition matrices associated with dive type 2 ($\hat \theta^{*(2)}$ and $\hat \Gamma^{*(2)}$) being especially poor estimates (see Table 9 and Figure 33 of the supplementary material).

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model is unable to estimate dive types as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by approximately five seconds. See Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive types and subdive states for each model.

>>>>

%\section{Simulation Study}
We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,t^*}$ is set to a scalar rather than a three dimensional vector. We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information under that model.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2T^*_t$ seconds of each dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $Z^*_{t,t^*} = \left\{\Zone_{t,t^*}, \Ztwo_{t,t^*}\right\}$ as defined in Equation (\ref{eqn:z}). Recall that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

The two simulated dive types differ in that dives of type one are much shorter on average (20 seconds) than dives of type two (80 seconds). The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher for subdive state 2 than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data. All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $X^*_{t,t^*}$ is the unknown subdive state associated with a particular model and $x^*_{t,t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. The average dive decoding accuracy is above 0.9 while the average subdive decoding accuracy is $1-10^{-12}$. All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases. In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are as severe or less severe than the other three models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to slightly underestimate the empirical standard errors. See Tables 6 through 9 of the supplementary material for detailed results.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its average dive and subdive decoding accuracy as well as its parameter estimate biases. However, unlike the CarHHMM-DFT, the HHMM-DFT greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material). These results suggest that the estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored. This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.

The CarHHMM is the worst-performing model in terms of accuracy, as its average dive decoding accuracy and average subdive decoding accuracy are both below $0.9$. This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct difference between the subdive states. The CarHHMM is much more likely to misclassify subdive state 1 as subdive state 2 than vice-versa (see Table \ref{table:accuracy}). In addition to its poor average subdive and dive decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters. Estimates associated with dive type 2 ($\theta^{(2)}$ and $\hat \Gamma^{*(2)}$) and subdive type 2 ($\theta^{*(\cdot,2)}$) are especially poor, likely due to the fact that the CarHHMM often classifies dive type 1 as dive type 2 and subdive state 1 as subdive state 2. See Section 2 of the supplementary material for more detailed results.

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of average subdive decoding accuracy, parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model does not estimate dive type as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a single Gamma distribution over the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and on average the value of $\hat \sigma$ is approximately 5 seconds less than the average sample standard deviation of dive duration, indicating significant model misspecification. See Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive types and subdive states for each model.

### 2020-10-30 Report/main.tex
commit before adding t_tilde

% Title, authors, affiliations
%\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
\title[]{Modelling state-switching behaviour in functional data with hidden Markov models}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Association parameters\sep clustered data\sep mean parameters\sep missing data\sep pairwise
likelihood\sep repeated measurements.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62???\sep secondary 62???}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
%Insert your abstract here; it should
%typically be up to ten lines long. Avoid symbols as much as possible. Formulas
%are strongly discouraged, and citations should be avoided. 
%The title and the abstract should be concise and descriptive;
%list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
%http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.

Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields. These data often exhibit complicated temporal dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA). Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs.
We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at fine scales. 
%We also provide several examples of viable fine-scale models.
As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. 
Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.

@@

% Email address for corresponding author
%\correspondingauthor[*]{\\\email{}} % only put this in after acceptance

%%%%%%%%%%%%
%\vfill\eject
%{\centerline{\Large{Notes from Nancy}}}
%\input{Sections/0627NH}
%\vfill\eject
%%%%%%%%%%%%%
%
\section{INTRODUCTION}
\input Sections/intro.tex
%
\section{MODELS AND PARAMETER ESTIMATION}
\label{sec:models}
\input Sections/models.tex
%
\section{KILLER WHALE CASE STUDY}
\label{sec:data}
\input Sections/results.tex
%
\section{SIMULATION STUDY}
\label{sec:sim_study}
\input Sections/simulation.tex
%
%\section{Case Study Results}
%\label{sec:case_study}
%\input Sections/results.tex
%%
\section{DISCUSSION}
\input Sections/discussion.tex
%
%\newpage
%\input Sections/bibliography.tex
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{FIGURES AND TABLES}
\input Sections/figures.tex
%
%\CJShistory

>>>>

% Title, authors, affiliations
%\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
\title[]{Modelling state-switching functional data with hidden Markov models}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Time series modelling\sep multi-state modelling\sep hierarchical modelling\sep biologging\sep statistical ecology\sep animal movement modelling.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
%Insert your abstract here; it should
%typically be up to ten lines long. Avoid symbols as much as possible. Formulas
%are strongly discouraged, and citations should be avoided. 
%The title and the abstract should be concise and descriptive;
%list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
%http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.

\iffalse
Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields. These data often exhibit complicated temporal dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA). Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs. We detail a general hierarchical approach based on HMMs which incorporates a wider class of models at fine scales. As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.
\fi

Sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA). We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM). The distribution of each curve is then defined by another general fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-correlation. As a case study, we use this framework to model the movement of a killer whale off the western coast of Canada. Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines. This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.

@@

% Email address for corresponding author
%\correspondingauthor[*]{\\\email{}} % only put this in after acceptance
%
\section{INTRODUCTION}
\input Sections/intro.tex
%
\section{MODELS AND PARAMETER ESTIMATION}
\label{sec:models}
\input Sections/models.tex
%
\section{KILLER WHALE CASE STUDY}
\label{sec:data}
\input Sections/results.tex
%
\section{SIMULATION STUDY}
\label{sec:sim_study}
\input Sections/simulation.tex
%
\section{DISCUSSION}
\input Sections/discussion.tex
%
%\newpage
%\input Sections/bibliography.tex
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{FIGURES AND TABLES}
\input Sections/figures.tex
%
%\CJShistory
